---
title: "Simulation for microbiome power analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
  pdf_document: default
vignette: >
  %\VignetteIndexEntry{Simulation for microbiome power analysis}
  %\VignettePackage{power.nb}
  %\VignetteEngine{knitr::rmarkdown}
---



## Introduction 


**Installation**

Install the package from GitHub using `devtools`:

```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("magronah/power.nb")
```

### Overview

This package allows you to:

- Fit statistical models to microbiome count data.

- Simulate realistic count datasets with different effect sizes.

- Test for differential abundance.

- Estimate power per taxon.

- Explore sample size and effect size sensitivity.


## Datasets 
We used the datasets by Dr. Nearing used in the paper ... cite{}. 

### Data simulation for power estimation

We would need to simulate microbiome data in order to estimate statistical power at the level of individual taxon.  The simulation framework implemented in `power.nb` is Mixture of Gaussian Mixture Model proposed in the paper by [Agronah and Bolker](https://doi.org/10.1371/journal.pone.0318820). There are one of two ways to obtain the parameters of the Mixture of Gaussian Mixture Model for data simulation. 

1) Users can either pre-specify a set of parameters 
**OR** 
2)  Estimating these parameters from existing datasets 


#### Prespecifing parameters for the Gaussian model


#### Estimating Gaussian parameters from existing data

The following steps outlines the process for estimating parameters if you have a dataset. To ensure that results are applicable to the study you intend conducting. 

similar to the one you intend using for your study. 
Use your microbiome data to fit a Gaussian mixture model:


##### Data preprocessing 
Low abundant tax exhibit high variability, posing challenges in detecting significant differences between groups cite{m}. Pre-filtering steps, as is routinely done in differential abundance analysis, is used to filter out these low abundant taxa. In this vignette, we filter out taxa with less that 10 count in less that 5 samples. 

```{r, eval=FALSE}
# load library
library(power.nb) 

#filter out low abundant taxa
filter_data  = filter_low_count(countdata = data,
                                metadata  = metadata,
                                abund_thresh = 10,
                                sample_thresh = 5,
                                sample_colname = "SampleID",
                                group_colname  = "Groups")

head(filter_data)
```

##### Model Fitting

```{r, eval=FALSE}
#Fit log mean count
logmean    =  log(rowMeans(filter_data))
logmeanFit =  logmean_fit(logmean, sig = 0.05, 
                         max.comp = 4, max.boot = 100)
names(logmeanFit)

#Estimate foldchange from Deseq
foldchange_est <- deseqfun(countdata = data,
                           metadata  = metadata,
                          alpha_level = 0.1,
                          ref_name  = "Control",
                          group_colname = "Groups",
                          sample_colname = "SampleID")

logfoldchange =  foldchange_est$log2FoldChange

## Fit foldchange from Deseq
logfoldchangeFit <- logfoldchange_fit(logmean,
                                       logfoldchange,
                                       ncore = 2,
                                       max_sd_ord = 2,
                                       max_np = 5,
                                       minval = -5,
                                       maxval = 5,
                                       itermax = 100,
                                       NP = 800,
                                       seed = 100)

logfoldchangeFit
```


## Simulating Data

- Simulate log mean count from the fitted Mixture of Gaussian (using existing dataset) or from the pre-specified parameters for the Gaussian Mixture 

- Simulate log fold change using the estimated parameters or the prespecified parameters and the simulated as the input for the variance and mean of the log fold change 

- Simulate dispersion parameters from the estimated nonlinear function of the dispersion function 

- Simulate count data from a negative binomial model and then 

- Use the simulated procedure outline above to simulate n number of count data 


```{r, eval=FALSE}
sim <- simulate_data(fit,
                     n_taxa = 1000,
                     n_per_group = c(30, 30),
                     fold_changes = c(1.5, 2, 4),
                     nsim = 100)
```


## Power estimation procedure

- Fit the count data with the negative binomial model in the desert package to estimate pvalues for each fold change estimated fold change

- Construct a vector of binaries with 1 for a taxa where value is less than a specified threshold and 0 otherwise. 

- We fit a Generalised additive model to predict the probability of obtaining 1  (also the statistical power) for any combination of mean count and log fold change 

- Figure shows a contour plot for 3  of the datasets. From the plots, there is low statistical power. 
  a Concatenation all the 



```{r, eval=FALSE}
####

```


## Sample Size Estimation


```{r, eval=FALSE}
####

```


## Appendix 

```{r, echo=FALSE}
library(knitr)
library(kableExtra)

# Example data (two rows like in your LaTeX version)
df <- data.frame(
  `Data description` = c("Example row 1", "Example row 2"),
  n1 = c(5, 5),
  mu1 = c("(2,3,4,5,7)", "(2,3,4,5,7)"),
  sigma1 = c("(2,3,4,5,7)", "(2,3,4,5,7)"),
  p1 = c("(2,3,4,5,7)", "(2,3,4,5,7)"),
  n2 = c(5, 5),
  mu2 = c("(2,3,4,5,7)", "(2,3,4,5,7)"),
  sigma2 = c("(2,3,4,5,7)", "(2,3,4,5,7)"),
  p2 = c("(2,3,4,5,7)", "(2,3,4,5,7)"),
  fx = c("(2,3,4,5,7)", "(2,3,4,5,7)")
)

kbl(df, booktabs = TRUE, longtable = TRUE,
    caption = "Data description and parameter estimates from actual microbiome datasets") %>%
  kable_styling(latex_options = c("hold_position", "repeat_header")) %>%
  add_header_above(c(" " = 1, "Log mean count" = 4, "Log foldchange" = 5)) %>%
  add_header_above(c(" " = 1, "Parameters" = 9)) # optional extra grouping

```


