---
title: "Power and Sample Size Estimation for Microbiome Analysis"
date: "`r Sys.Date()`"
author: "Michael Agronah and Benjamin Bolker"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
  pdf_document: default
vignette: >
  %\VignetteIndexEntry{Simulation for microbiome power analysis}
  %\VignettePackage{power.nb}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib
---


# Acknowledgments

We would like to express our sincere gratitude to Dr. Jacob T. Nearing for his generosity in granting us access to the 38 microbiome datasets used in the paper “Microbiome differential abundance methods produce different results across 38 datasets” [@Nearing2022]. These datasets were invaluable in developing, testing, and validating this R package for power and sample size calculation in microbiome studies.

#  Package Description and Functionalities 

`power.nb` is a package developed for estimating statistical power and sample sizes in differential abundance microbiome studies. The package presents novel methods for estimating statistical power for individual taxa (feature, otu, or species) and for sample size calculation accounting for effect sizes and mean abundance of taxa.


`power.nb` also presents two novel microbiome data simulations frameworks: `MixGassSim` and `RRSim`. The method implemented in `MixGassSim` models the distribution of mean abundance of taxa and the distribution of fold change (as a function of mean abundance of taxa) by finite Mixture of Gaussian distributions.  Microbiome count data is then simulated from a negative binomial distribution. Detailed description of the `MixGassSim` method is presented in the paper ["Investigating statistical power of differential abundance studies"](https://doi.org/10.1371/journal.pone.0318820) [@agronahnbolker].  

`RRSim` (this simulation method is still under development and some functions in `RRSim` are yet to tested), on the other hand models the mean abundance of all taxa jointly with a mixed-effects model while accounting for correlations among taxa within subjects and zero inflation in microbiome data. Given the high dimensionality of microbiome data (usually having hundreds or thousands of taxa), modelling correlations between taxa requires estimating thousands or even millions of parameters for the correlation matrix. `RRSim` uses the reduced rank functionality implemented in the `glmmTMB` [@magnusson2017package] packages to reduce the number of parameter estimates required of the variance-correlation matrix. Details of this method is presented in the thesis ["A Novel Approach for Simulation-Based Power Estimation and Joint Modeling of Microbiome Counts"](https://macsphere.mcmaster.ca/handle/11375/31784) [@agronah2025novel]

#  Installation

Install the package from GitHub using `devtools`:

```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("magronah/power.nb")
```

# Simulating microbiome data

Since the goal of differential abundance studies is to identify taxa that differ significantly between groups, statistical power must be estimated at the level of individual taxa. Because of the complexity of microbiome data, analytical approaches based on theoretical distributions of test statistics (e.g., non-central t or chi-squared distributions) are not feasible [@arnold2011simulation]. Therefore, power estimation typically relies on simulation-based methods that can mimic the characteristics of real microbiome data.

Statistical power for a given taxon depends on both its mean abundance and effect size (defined in this package as fold changes) [@agronahnbolker]. Thus, estimating the distributions of mean abundance and fold change of taxa explicitly will offer great benefit for estimation of statistical power for individual taxon. We use the `MixGaussSim` simulation approach to simulate microbiome data for power calculation due to its flexibility in modelling the distributions of taxa mean abundances and effect sizes explicitly as as well as their relationship. 


## Two ways to obtain parameters for data simulation using MixGaussSim
There are one of two ways to obtain the parameters of the `MixGaussSim` method for data simulation.  Users can:

**1. pre-specify a set of parameters:** To help users decide on plausible parameter values, we have run  `MixGaussSim` on some of the microbiome datasets used in the paper ["Microbiome differential abundance methods produce different results across 38 datasets"](10.1038/s41467-022-28034-z) [@Nearing2022] and presented the parameter estimates  obtained from each of these datasets ([see Appendix for the parameter estimates from these datasets](#appendix) ). 

**2. estimate parameters from an existing dataset:** Secondly, users can also obtain estimates directly from fitting the model implemented in  `MixGaussSim`  to existing datasets. The following steps outlines the process for estimating parameters from an existing dataset. 

## Dataset

We use a subset of the Obsesity dataset, one of the datasets analyzed in [@Nearing2022]. The full dataset contains 52 samples and 1203 taxa.

```{r,message=FALSE}
##Load libraries
library(tidyverse)
library(dplyr)
library(power.nb)
library(patchwork)

data_path =  "~/Documents/power.nb/data"
##Read count data and metadata
data <- read.table(
  file.path(data_path, "ob_ross_ASVs_table.tsv"),
  header = TRUE, sep = "\t",
  check.names = FALSE, comment.char = "", 
  row.names = 1
)

metadata <- read.table(
  file.path(data_path, "ob_ross_metadata.tsv"),
  header = TRUE, sep = "\t",
  check.names = FALSE, comment.char = ""
)

metadata <- metadata %>%
  setNames(c("SampleID", "Groups"))

dim(data); dim(metadata)
head(data); head(metadata)

```

## Pre-filtering low abundant taxa

Low abundant tax exhibit high variability, posing challenges in detecting significant differences between groups [@love2014moderated]. Pre-filtering steps, as is routinely done in differential abundance analysis, is used to filter out these low abundant taxa. We filter out taxa with less that 10 count in less that 5 samples. 

```{r}
filter_data  =  filter_low_count(countdata     =  data,
                                metadata       =  metadata,
                                abund_thresh   =  10,
                                sample_thresh  =  5,
                                sample_colname =  "SampleID",
                                group_colname  =  "Groups")

dim(filter_data)
```

## Foldchange and Dispersion Estimation

We estimate fold change for each taxon and dispersion parameters using the negative binomial model implemented in the `DESeq2` package [@love2014moderated]. The negative binomial model is widely used in both microbiome and RNA-seq analyses and is implemented in the `NBZIMM` [@yi2020nbzimm] and `edgeR` R packages [@robinson2010edger].  The model implemented in the `DESeq2` package is described as follows:

Let $K_{ij}$ represent the observed count for the $i^{\text{th}}$ taxon in the $j^{\text{th}}$ sample. The model assumes that  
$$K_{ij} \sim \text{NB}(\mu_{ij}, \alpha_{ij}),$$
where the mean is defined as $\mu_{ij} = s_j q_{ij}$, and $s_j$ is the normalization factor for sample $j$. The expected abundance prior to normalization, $q_{ij}$, is modeled as  
$$\log q_{ij} = \sum_r x_{jr} \beta_{ir},$$
with $x_{jr}$ denoting the covariates and $\beta_{ir}$ the associated coefficients. The dispersion parameter is assumed constant across samples for each taxon, such that $\alpha_{ij} = \alpha_i$. This formulation links the variance and mean through  
$$\mathrm{Var}(K_{ij}) = \mu_i + \alpha_i \mu_i^2.$$

Estimation of the regression coefficients $\hat{\beta}_{ir}$ and dispersion parameters $\hat{\alpha}_i$ was carried out using the empirical Bayes shrinkage procedures implemented in the `DESeq2` package [@love2014moderated,@anders2010differential]. 


```{r,message=FALSE}

unique(metadata$Groups)

foldchange_est <- deseqfun(countdata      =   filter_data,
                           metadata       =   metadata,
                           alpha_level    =   0.1,
                           ref_name       =  "H",
                           group_colname  =  "Groups",
                           sample_colname =  "SampleID")

logfoldchange =  foldchange_est$deseq_estimate$log2FoldChange

head(logfoldchange)
```


## Fit  log mean count 

We modelled the log mean abundance (defined as the logarithm of the arithmetic mean abundance across both control and treatment groups) as a finite mixture of Gaussian distributions. To determine the optimal number of mixture components, denoted by $k$, we employed a parametric bootstrap approach to sequentially compare models with $k$ and $k+1$ components. 

Let the observed data be denoted by $\mathbf{y} = (y_1, y_2, \dots, y_n)$, where each $y_i$ represents the log mean abundance for the $i^{\text{th}}$ taxon. The mixture model with $k$ components is given by

$$f_k(y_i \mid \boldsymbol{\theta}_k) = \sum_{j=1}^{k} \pi_{j,k} \, \phi(y_i \mid \mu_{j,k}, \sigma_{j,k}^2),$$

where $\pi_{j,k}$ are the mixing proportions satisfying $\sum_{j=1}^{k} \pi_{j,k} = 1$, and $\phi(\cdot \mid \mu, \sigma^2)$ denotes the normal density with mean $\mu$ and variance $\sigma^2$.

To test whether adding an additional component improves model fit, we formulated the hypotheses as:

$$\begin{cases} H_0: \mathbf{y} \sim f_k(y_i \mid \boldsymbol{\theta}_k) & \text{(the data follow a $k$-component mixture model)} \\[4pt]
H_a: \mathbf{y} \sim f_{k+1}(y_i \mid \boldsymbol{\theta}_{k+1}) & \text{(the data follow a $(k+1)$-component mixture model)} \end{cases}$$

The test statistic is the likelihood ratio statistic:

$$\Lambda = 2 \left[ \ell_{k+1}(\widehat{\boldsymbol{\theta}}_{k+1}) - \ell_{k}(\widehat{\boldsymbol{\theta}}_{k}) \right],$$
where $\ell_{k}(\widehat{\boldsymbol{\theta}}_{k})$ and $\ell_{k+1}(\widehat{\boldsymbol{\theta}}_{k+1})$ are the maximized log-likelihoods under the $k$- and $(k+1)$-component models, respectively.

To approximate the null distribution of $\Lambda$, we performed a parametric bootstrap using $B = 100$ bootstrap samples generated under $H_0$. For each bootstrap replicate $b = 1, \dots, B$, data were simulated from the fitted $k$-component model, and both the null and alternative models were re-fitted to obtain $\Lambda^{(b)}$. The empirical $p$-value is computed as

$$p = \frac{1}{B} \sum_{b=1}^{B} I\left( \Lambda^{(b)} \ge \Lambda_{\text{obs}} \right),$$
where $\Lambda_{\text{obs}}$ is the observed likelihood ratio statistic. 

If $p < 0.05$, the null hypothesis $H_0$ is rejected in favor of the $(k+1)$-component model. Testing proceeds sequentially for $k = 1, 2, \dots, 5$ until the $p$-value exceeds the $5\%$ significance threshold, at which point the $k$-component model from the last accepted test is selected as the optimal mixture size [@benaglia2010mixtools].

```{r, message=FALSE, warning=FALSE}
logmean    =  log(rowMeans(filter_data))

quiet <- function(expr) {
  out <- suppressWarnings(suppressMessages(
    capture.output(res <- eval.parent(substitute(expr)))
  ))
  res}

logmeanFit =   quiet(logmean_fit(logmean, sig = 0.05,
                                 max.comp = 4, max.boot = 100))

logmeanFit
```



## Fit  Dispersion estimates

Because dispersion tends to depend on mean count abundance with rarer taxa showing greater variability we modeled the dispersion as a nonlinear function of the mean, following the approach implemented in `DESeq2`. The nonlinear function is defined by

<span id="disp-eq"></span>
$$
d = c_0 + \frac{c_1}{m}
$$

where $d$ and $m$ denote the scaled dispersion and mean abundance respectively. The term $c_0$ represents the asymptotic dispersion level for high abundance taxa, and $c_1$ captures additional dispersion variability. This procedure enables estimation of dispersion values corresponding to given mean counts, facilitating their use in downstream simulations and power analyses.


```{r,message=FALSE, warning=FALSE}
dispersion    =  foldchange_est$dispersion
dispersionFit =  dispersion_fit(dispersion, logmean)
dispersionFit$param
```

## Fit log fold change

We also modelled the log fold change as a finite mixture of Gaussian distributions. Fold change is typically related to mean abundance [@love2014moderated]; therefore, we expressed the mean and standard deviation parameters of the individual Gaussian components as functions of log mean abundance. Detailed description model fitted to log fold changes is presented in the paper ["Investigating statistical power of differential abundance studies"](https://doi.org/10.1371/journal.pone.0318820) [@agronahnbolker].  


```{r, eval = FALSE}
## Fit foldchange from Deseq
logfoldchangeFit <- logfoldchange_fit(logmean,
                                       logfoldchange,
                                       ncore = 3,
                                       max_sd_ord = 2,
                                       max_np = 5,
                                       minval = -5,
                                       maxval = 5,
                                       itermax = 100,
                                       NP = 800,
                                       seed = 100)

head(logfoldchangeFit)
```


## Simulate count microbiome data

To simulate microbiome data, we simulate log mean count and corresponding log fold change. Using the simulated mean count and fold change, we compute the mean count of taxa for each group (control and treatment groups) and estimate dispersion values for each sample using the fitted nonlinear function in the equation [here](#disp-eq).


### Simulate log mean count and log fold change

```{r, eval = FALSE}

### Simulated log mean count and log foldchange using the fitted parameters

logmean_param       =  logmeanFit$logmean_param
logfoldchange_param =  logfoldchangeFit

notu  = dim(filter_data)[1]
logmean_sim  =  logmean_sim_fun(logmean_param, notu)

logfoldchange_sim  =  logfoldchange_sim_fun(logmean_sim = logmean_sim,
                                            logfoldchange_param = 
                                            logfoldchange_param,
                                            max_lfc  = 15,
                                            max_iter = 30000)
```

### Simulate counts from the negative binomial distribution 

The dispersion values estimated internally with the `countdata_sim_fun` function. 

```{r, eval = FALSE}

ntreat = sum(metadata$Groups == "OB")
ncont  = sum(metadata$Groups == "H")
dispersion_param  =  dispersionFit$param

nsim = 100
countdata_sims = countdata_sim_fun(logmean_param,
                                   logfoldchange_param,
                                   dispersion_param,
                                   nsamp_per_group = NULL,
                                   ncont = ncont,
                                   ntreat = ntreat,
                                   notu,
                                   nsim = nsim,
                                   disp_scale = 0.3,
                                   max_lfc = 15,
                                   maxlfc_iter = 1000,
                                   seed = 121)


dim(countdata_sims$treat_countdata_list$sim_1)
dim(countdata_sims$control_countdata_list$sim_1)
dim(countdata_sims$countdata_list$sim_1)
dim(countdata_sims$metadata_list$sim_1)


length(countdata_sims$logfoldchange_list$sim_1)
length(countdata_sims$logmean_list$sim_1)
###############################################################
```

### Comparing the distributions of mean count and fold change from simulation with actual data

We compare the following comparisions bettween our simulation and the actual dataset 

- we come distribution of variances of counts of taxa and 

- the means of taxa from out simulation with those of the actual dataset

```{r, eval = FALSE}
dd  = data.frame(logmean  = c(logmean,
                              countdata_sims$logmean_list$sim_1),
                 logfoldchange  = c(foldchange_est$logfoldchange,
                                  countdata_sims$logfoldchange_list$sim_1),
                 type  = rep(c("observation","simulation"), each = notu) )

p1 = ggplot(dd, aes(x = logmean, group = type, colour = type)) +
  geom_density()  +
  theme_bw()

p2 = ggplot(dd, aes(x = logfoldchange, group = type, colour = type)) +
  geom_density()  +
  theme_bw()

(p1|p2) + plot_layout(guides = "collect")
```


# Power estimation fir individual taxa proceed

## Simulating microbiome datasets

## Estimate p-values associated with the simulated fold changes for each taxa

We now run each simulated count adata to the to calculate pvalues associated to each  foldchanes for each taxa. 


```{r, eval = FALSE}

###############################################################
countdata_list  =   countdata_sims$countdata_list
metadata_list   =   countdata_sims$metadata_list
desq_est   =   deseq_fun_est(metadata_list =  metadata_list,
                            countdata_list =  countdata_list,
                            alpha_level    =  0.1,
                            group_colname  = "Groups",
                            sample_colname = "Samples",
                            num_cores      =  4,
                            ref_name       = "control")
###############################################################
true_lmean_list    =    countdata_sims$logmean_list
deseq_est_list     =    lapply(desq_est, function(x){x$deseq_estimate})
true_lfoldchange_list = countdata_sims$logfoldchange_list
###############################################################
```

## Fit Generalized Additive Model (GAM) for power estimation 

```{r, eval = FALSE}
gamFit <- gam_fit(deseq_est_list,
                  true_lfoldchange_list,
                  true_lmean_list,
                  grid_len = 50,
                  alpha_level=0.1)
  

cont_breaks     =  seq(0,1,0.1)
combined_data   =  gamFit$combined_data
power_estimate  =  gamFit$power_estimate
#plot(gamFit$fit_2d)

contour_plot <- contour_plot_fun(combined_data,
                                 power_estimate,
                                 cont_breaks)
contour_plot
```


## Predict statistical power given fold change and mean abundance for taxa



# Sample size calculation

## Simulate count data for various ranges of sample sizes

```{r, eval = FALSE}
nsim = 50
nsamp_vec = c(20, 50, 70)
countdata_sims_list  =  list()
for(j in 1:length(nsamp_vec)){
  countdata_sims_list[[j]]  =  countdata_sim_fun(logmean_param,
                                                 logfoldchange_param,
                                                 dispersion_param,
                                                 nsamp_per_group = nsamp_vec[j],
                                                 ncont  = NULL,
                                                 ntreat = NULL,
                                                 notu,
                                                 nsim = nsim,
                                                 disp_scale = 0.3,
                                                 max_lfc = 15,
                                                 maxlfc_iter = 1000,
                                                 seed = 121)
}

names(countdata_sims_list) = paste0("sample_",nsamp_vec)
###############################################################
```

## Estimate p-values associated to fold changes for each taxa for simulated data per sample size

```{r, eval = FALSE}
desq_est_list  =  list()
for(i in 1:length(countdata_sims_list)){
  
  countdata_list       =   countdata_sims_list[[i]]$countdata_list
  metadata_list        =   countdata_sims_list[[i]]$metadata_list
  desq_est_list[[i]]   =   deseq_fun_est(metadata_list =  metadata_list,
                               countdata_list =  countdata_list,
                               alpha_level    =  0.1,
                               group_colname  = "Groups",
                               sample_colname = "Samples",
                               num_cores      =  4,
                               ref_name       = "control")
  
}
```


## Fit Generalized Additive Model (GAM) for power estimation 



## Estimate sample size for a given statistical power, fold change and mean abundance




##### Data preprocessing 
Low abundant tax exhibit high variability, posing challenges in detecting significant differences between groups cite{m}. Pre-filtering steps, as is routinely done in differential abundance analysis, is used to filter out these low abundant taxa. In this vignette, we filter out taxa with less that 10 count in less that 5 samples. 

```{r, eval=FALSE}
# load library
library(power.nb) 

#filter out low abundant taxa
filter_data  = filter_low_count(countdata = data,
                                metadata  = metadata,
                                abund_thresh = 10,
                                sample_thresh = 5,
                                sample_colname = "SampleID",
                                group_colname  = "Groups")

head(filter_data)
```

##### Model Fitting

```{r, eval=FALSE}
#Fit log mean count
logmean    =  log(rowMeans(filter_data))
logmeanFit =  logmean_fit(logmean, sig = 0.05, 
                         max.comp = 4, max.boot = 100)
names(logmeanFit)

#Estimate foldchange from Deseq
foldchange_est <- deseqfun(countdata = data,
                           metadata  = metadata,
                          alpha_level = 0.1,
                          ref_name  = "Control",
                          group_colname = "Groups",
                          sample_colname = "SampleID")

logfoldchange =  foldchange_est$log2FoldChange

## Fit foldchange from Deseq
logfoldchangeFit <- logfoldchange_fit(logmean,
                                       logfoldchange,
                                       ncore = 2,
                                       max_sd_ord = 2,
                                       max_np = 5,
                                       minval = -5,
                                       maxval = 5,
                                       itermax = 100,
                                       NP = 800,
                                       seed = 100)

logfoldchangeFit
```


## Simulating Data

- Simulate log mean count from the fitted Mixture of Gaussian (using existing dataset) or from the pre-specified parameters for the Gaussian Mixture 

- Simulate log fold change using the estimated parameters or the prespecified parameters and the simulated as the input for the variance and mean of the log fold change 

- Simulate dispersion parameters from the estimated nonlinear function of the dispersion function 

- Simulate count data from a negative binomial model and then 

- Use the simulated procedure outline above to simulate n number of count data 


```{r, eval=FALSE}
sim <- simulate_data(fit,
                     n_taxa = 1000,
                     n_per_group = c(30, 30),
                     fold_changes = c(1.5, 2, 4),
                     nsim = 100)
```


## Power estimation procedure

- Fit the count data with the negative binomial model in the desert package to estimate pvalues for each fold change estimated fold change

- Construct a vector of binaries with 1 for a taxa where value is less than a specified threshold and 0 otherwise. 

- We fit a Generalised additive model to predict the probability of obtaining 1  (also the statistical power) for any combination of mean count and log fold change 

- Figure shows a contour plot for 3  of the datasets. From the plots, there is low statistical power. 
  a Concatenation all the 







# Appendix 

```{r, echo=FALSE}
library(knitr)
library(kableExtra)

# Example data (two rows like in your LaTeX version)
df <- data.frame(
  `Data description` = c("Example row 1", "Example row 2"),
  n1 = c(5, 5),
  mu1 = c("(2,3,4,5,7)", "(2,3,4,5,7)"),
  sigma1 = c("(2,3,4,5,7)", "(2,3,4,5,7)"),
  p1 = c("(2,3,4,5,7)", "(2,3,4,5,7)"),
  n2 = c(5, 5),
  mu2 = c("(2,3,4,5,7)", "(2,3,4,5,7)"),
  sigma2 = c("(2,3,4,5,7)", "(2,3,4,5,7)"),
  p2 = c("(2,3,4,5,7)", "(2,3,4,5,7)"),
  fx = c("(2,3,4,5,7)", "(2,3,4,5,7)")
)

kbl(df, booktabs = TRUE, longtable = TRUE,
    caption = "Data description and parameter estimates from actual microbiome datasets") %>%
  kable_styling(latex_options = c("hold_position", "repeat_header")) %>%
  add_header_above(c(" " = 1, "Log mean count" = 4, "Log foldchange" = 5)) %>%
  add_header_above(c(" " = 1, "Parameters" = 9)) # optional extra grouping

```


